### 2015-08-06 background.tex
Complete pass.

The most common method for avoiding races is by protecting sets of program statements that access a shared resource with \emph{locks}, forming \emph{critical sections}.  Fig.~\ref{fig:lock_example} shows how to use locking to eliminate the races in Fig.~\ref{fig:data_race_example}. Because the \texttt{return} statement can potentially race on the \texttt{f\_pos} field, we store the result in a temporary variable \texttt{res} inside the critical section.
%
Careless use of locks has many well-known pitfalls~\cite{sutter2005software}: coarse-grained locking can hurt performance as it limits the opportunity for concurrency, while fine-grained locking can easily lead to deadlocks. Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite[p.\ 123]{corbet2005linux}, such as atomic read-modify-write operations, in this work we focus on locks as they are widely used.\footnote{We treat lock-free operations soundly by regarding them as not providing any protection between threads; this can lead to false alarms.}

@@

Lockset analysis for a concurrent program starts by creating a \emph{current} lockset $\mathit{CLS}_T$ for each thread $T$ of the program, and a lockset $\mathit{LS}_s$ for each shared variable $s$ used in the program. Initially, every $\mathit{CLS}_T$ is empty because the threads do not hold any locks on program start. In addition, every $\mathit{LS}_s$ is initialized to the set of all locks manipulated by the program since initially each access to $s$ is (vacuously) protected by every lock. The program is executed as usual (with threads scheduled according to the OS scheduler), except that instrumentation is added to update locksets as follows.
%
After each \emph{lock} and \emph{unlock} operation by $T$, $\mathit{CLS}_T$ is updated to reflect the locks currently held by $T$.
%
When $T$ accesses a shared variable $s$, $\mathit{LS}_s$ is updated to the intersection of $\mathit{LS}_s$ with $\mathit{CLS}_T$, which removes any locks that are not common to the two locksets.
%
If $\mathit{LS}_s$ becomes empty as a result, a warning is issued that the access to $s$ may be unprotected.

@@

In contrast to more sophisticated race analyses that encode a \emph{happens-before} relation between threads~\cite{lamport1978time} (e.g.\ using vector clocks), lockset analysis is easy to implement, lightweight, and thus has the potential to scale well.  The technique, though, suffers from imprecision (i.e.\ can report false bugs) since a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}. Furthermore, the code coverage in dynamic lockset analyzers, such as Eraser, is limited by the execution paths that are explored under a given scheduler. Since the tool runs the program in a truly concurrent environment, it only gets to see what the thread did on the particular schedule that was followed.

>>>>

The most common method for avoiding races is by protecting sets of program statements that access a shared resource with \emph{locks}, forming \emph{critical sections}.  Fig.~\ref{fig:lock_example} shows how to use locking to eliminate the races in Fig.~\ref{fig:data_race_example}. \COULDCUT{Because the \texttt{return} statement can potentially race on the \texttt{f\_pos} field, we store the result in a temporary variable \texttt{res} inside the critical section.}
%
Careless use of locks has many well-known pitfalls~\cite{sutter2005software}: coarse-grained locking can hurt performance as it limits the opportunity for concurrency, while fine-grained locking can easily lead to deadlocks. Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite[p.\ 123]{corbet2005linux}, such as atomic read-modify-write operations, in this work we focus on locks as they are widely used.\footnote{We treat lock-free operations soundly by regarding them as not providing any protection between threads; this can lead to false alarms.}

@@

Lockset analysis for a concurrent program starts by creating a \emph{current} lockset $\mathit{CLS}_T$ for each thread $T$ of the program, and a lockset $\mathit{LS}_s$ for each shared variable $s$ used in the program. Initially, every $\mathit{CLS}_T$ is empty because the threads do not hold any locks initially. In addition, every $\mathit{LS}_s$ is initialized to the set of all locks manipulated by the program since initially each access to $s$ is (vacuously) protected by every lock. The program is executed as usual (with threads scheduled according to the OS scheduler), except that instrumentation is added to update locksets as follows.
%
After each \emph{lock} and \emph{unlock} operation by $T$, $\mathit{CLS}_T$ is updated to reflect the locks currently held by $T$.
%
When $T$ accesses a shared variable $s$, $\mathit{LS}_s$ is updated to the intersection of $\mathit{LS}_s$ with $\mathit{CLS}_T$, which removes any locks that are not common to the two locksets.
%
If $\mathit{LS}_s$ becomes empty as a result, a warning is issued that the access to $s$ may be unprotected.

@@

In contrast to more sophisticated race analyses that encode a \emph{happens-before} relation between threads~\cite{lamport1978time} (e.g.\ using vector clocks), lockset analysis is easy to implement, lightweight, and thus has the potential to scale well.  The technique, though, can report false alarms since a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}. Furthermore, the code coverage in dynamic lockset analyzers, such as Eraser, is limited by the execution paths that are explored under a given scheduler.

### 2015-08-06 conclusion.tex
Complete pass.

In this paper we presented \whoop, a new automated approach for detecting all possible data races in device drivers. Compared to traditional data race detection techniques that are based on happens-before and typically attempt to explore as many thread interleavings as possible (and thus face code coverage and scalability issues), \textsc{Whoop} uses over-approximation and symbolic pairwise lockset analysis, which scales well. Exploiting the race-freedom guarantees provided by \whoop, we showed that we can achieve a sound partial-order reduction that can significantly accelerate \corral, a state-of-the-art concurrency bug-finder. Combining \whoop and \corral, we analyzed \sizeOfBenchmarks drivers from the Linux 4.0 kernel, showing that our technique can analyze drivers 1.5--20$\times$ faster than standalone \corral.

>>>>

In this paper we presented \whoop, a new automated approach for detecting all possible data races in device drivers. Compared to traditional data race detection techniques that are based on happens-before and typically attempt to explore as many thread interleavings as possible (and thus face code coverage and scalability issues), \textsc{Whoop} uses over-approximation and symbolic pairwise lockset analysis, which scales well. Exploiting the race-freedom guarantees provided by \whoop, we showed that we can achieve a sound partial-order reduction that can significantly accelerate \corral, a state-of-the-art concurrency bug-finder. \COULDCUT{Combining \whoop and \corral, we analyzed \sizeOfBenchmarks drivers from the Linux 4.0 kernel, showing that our technique can analyze drivers 1.5--20$\times$ faster than standalone \corral.}

### 2015-08-06 corral.tex
Complete pass.

\corral is a symbolic bounded verifier for Boogie IVL that uses the Z3 SMT solver to statically reason about program behaviors. It checks for violations of provided assertions, and reports a precise counterexample if an assertion violation is found. \corral performs bounded exploration of a concurrent program in two steps. First, given a bound on the number of allowed context-switches, the concurrent program is appropriately \emph{sequentialized}, and the generated sequential version preserves reachable states of the original concurrent program~\cite{popl2011-eqr,cav2009-lqr,cavLalR08}. Then, \corral attempts to prove bounded (in terms of the number of loop iterations and recursion depth) sequential reachability of a bug in a goal-directed, lazy fashion to postpone state space explosion when analyzing a large program. It uses two key techniques to achieve this: (i) variable abstraction, where it attempts to identify a minimal set of shared variables that have to be precisely tracked in order to discharge all assertions; and (ii) stratified inlining, where it attempts to inline procedures on-demand as they are required for proving program assertions.

@@

\corral is inherently unsound (i.e.\ can miss real races), because it performs bounded verification. However, \corral is precise and, assuming a precise environmental model, it will only report true races. \whoop takes advantage of this precision to report only feasible data races. Also note that our instrumentation conveniently tolerates most benign races, as it does not report a write-write race if two write accesses update the same shared memory location with the same value.

@@

We have implemented two different \texttt{yield} instrumentations in \whoop: \yieldcoarse and \yieldmr.
%
The first instruments \texttt{yield} statements in a binary fashion: if \whoop proves an entry point pair as race free, then it will instrument a \texttt{yield} only after each lock and unlock statement of the pair; else if \whoop finds that a pair might race, then it will instrument a \texttt{yield} after all visible operations of the pair.
%
\yieldmr is a finer-grained instrumentation: it instruments a \texttt{yield} only after each access to a memory region that might race in the pair (regardless if the pair has not been fully proven as race-free), and after each lock and unlock. In our experiments (see \S\ref{evaluation}), \yieldmr is significantly faster than \yieldcoarse.

>>>>

\corral is a symbolic bounded verifier for Boogie IVL that uses the Z3 SMT solver to statically reason about program behaviors. It checks for violations of provided assertions, and reports a precise counterexample if an assertion violation is found. \corral performs bounded exploration of a concurrent program in two steps. First, given a bound on the number of allowed context-switches, the concurrent program is appropriately \emph{sequentialized}, and the generated sequential version preserves reachable states of the original concurrent program up to the context bound~\cite{popl2011-eqr,cav2009-lqr,cavLalR08}. Then, \corral attempts to prove bounded (in terms of the number of loop iterations and recursion depth) sequential reachability of a bug in a goal-directed, lazy fashion to postpone state space explosion when analyzing a large program. It uses two key techniques to achieve this: (i) variable abstraction, where it attempts to identify a minimal set of shared variables that have to be precisely tracked in order to discharge all assertions; and (ii) stratified inlining, where it attempts to inline procedures on-demand as they are required for proving program assertions.

@@

\corral is inherently unsound (i.e.\ can miss real races), because it performs bounded verification. However, \corral is precise and, assuming a precise environmental model, it will only report true races. \whoop takes advantage of this precision to report only feasible data races. Also note that our instrumentation conveniently tolerates some benign races, as it does not report a write-write race if two write accesses update the same shared memory location with the same value.

@@

We have implemented two different \texttt{yield} instrumentations in \whoop: \yieldcoarse and \yieldmr.
%
The first instruments \texttt{yield} statements in a binary fashion: if \whoop proves an entry point pair (EPP) as race-free, then it will instrument a \texttt{yield} only after each lock and unlock statement of the pair; else if \whoop finds that a pair might race, then it will instrument a \texttt{yield} after all visible operations of the pair.
%
\yieldmr is a finer-grained instrumentation: it instruments a \texttt{yield} only after each access to a memory region (MR) that might race in the pair (regardless if the pair has not been fully proven as race-free), and after each lock and unlock. In our experiments (see \S\ref{evaluation}), \yieldmr is significantly faster than \yieldcoarse.

### 2015-08-06 evaluation.tex
Complete pass.

We can see that \whoop reports more races than \corral does. This is because \whoop employs an over-approximating shared state abstraction to conservatively model the effects of additional threads when analyzing an entry point pair, and because lockset analysis is inherently imprecise; both factors can lead to false positives.  On the other hand, \corral is precise, but can miss races because only a limited number of context-switches are considered.  Another issue with \corral is loop coverage due to unsound loop unrolling. To tackle this, we enable the built-in loop over-approximation described in previous work~\cite{lal2014powering}. This can potentially lead \corral to report false bugs, but we have not seen this in practice. Furthermore, when we apply \corral to a pair of entry points, we just check the specific pair and do not account for the effects of other threads (see \S\ref{corral}); this can also cause \corral to miss some races. Standalone \corral did not discover any data races that \whoop did not already report.  This is expected, as \whoop aims for soundness, and increases our confidence in the implementation of \whoop.

Most of the races that \whoop and \corral discovered fall into two categories. The first is about accessing a global counter (or flag) from concurrent entry points, without holding a lock. This might be for performance, and indeed a lot of the races we found might be benign. Even benign races, though, lead to undefined behavior according to the C standard, and it is well known that undefined behaviors can lead to unexpected results when combined with aggressive compiler optimizations. The second is about an entry point modifying a field of a struct (either global or passed as a parameter) without holding a lock. This can lead to a race if another entry point simultaneously accesses the same field of the same struct.

>>>>

We can see that \whoop reports more races than \corral does. This is because \whoop employs an over-approximating shared state abstraction to conservatively model the effects of additional threads when analyzing an entry point pair, and because lockset analysis is inherently imprecise; both factors can lead to false positives.  On the other hand, \corral is precise, but can miss races because only a limited number of context-switches are considered.  Another issue with \corral is loop coverage due to unsound loop unrolling. To tackle this, we enable the built-in loop over-approximation described in previous work~\cite{lal2014powering}. This can potentially lead \corral to report false bugs, but we have not seen this in practice. Furthermore, when we apply \corral to a pair of entry points, we just check the specific pair and do not account for the effects of other threads (see \S\ref{corral}); this can also cause \corral to miss some races. Standalone \corral did not discover any data races that \whoop did not already report.  This is expected, as \whoop aims for soundness, and this increases our confidence in the implementation of \whoop.

Most of the races that \whoop and \corral discovered fall into two categories. The first is about accessing a global counter (or flag) from concurrent entry points, without holding a lock. This might be for performance, and indeed a lot of the races we found might be benign. Even benign races, though, lead to undefined behavior according to the C standard\COULDCUT{, and it is well known that undefined behaviors can lead to unexpected results when combined with aggressive compiler optimizations}. The second is about an entry point modifying a field of a struct (either global or passed as a parameter) without holding a lock. This can lead to a race if another entry point simultaneously accesses the same field of the same struct.

### 2015-08-06 intro.tex
Complete pass.

Device drivers are complex pieces of system-level software, operating at the thin boundary between hardware and software to provide an interface between the operating system and hardware devices that are attached to a computer. Drivers are notoriously hard to develop and debug~\cite{corbet2005linux}. This has a negative impact on hardware product releases, as time-to-market is commonly dominated by driver development, verification, and validation schedules~\cite{yavatkar2012era}.
%
Even after a driver has shipped, it typically has many undetected errors: Chou et al.~\cite{chou2001empirical} gathered data from 7 years of Linux kernel releases and found that the relative error-rate in driver source code is up to 10 times higher than in any other part of the kernel, while Swift et al.~\cite{Swift2003windowsxp} found that 85\% of the system crashes in Windows XP are due to faulty drivers. Regarding \emph{concurrency bugs}, a recent study~\cite{ryzhyk2009dingo} established that they account for 19\% of the total bugs in Linux drivers, showcasing their significance. The majority of these concurrency bugs were \emph{data races} or \emph{deadlocks} in various configuration functions and hot-plugging handlers.

>>>>

Device drivers are complex pieces of system-level software, operating at the thin boundary between hardware and software \COULDCUT{to provide an interface between the operating system and hardware devices that are attached to a computer}. Drivers are notoriously hard to develop and debug~\cite{corbet2005linux}. This has a negative impact on hardware product releases, as time-to-market is commonly dominated by driver development, verification, and validation schedules~\cite{yavatkar2012era}.
%
Even after a driver has shipped, it typically has many undetected errors: Chou et al.~\cite{chou2001empirical} gathered data from 7 years of Linux kernel releases and found that the relative error-rate in driver source code is up to 10 times higher than in any other part of the kernel, while Swift et al.~\cite{Swift2003windowsxp} found that 85\% of the system crashes in Windows XP are due to faulty drivers. Regarding \emph{concurrency bugs}, a recent study~\cite{ryzhyk2009dingo} established that they account for 19\% of the total bugs in Linux drivers, showcasing their significance. The majority of these concurrency bugs were \emph{data races} or \emph{deadlocks} in various configuration functions and hot-plugging handlers.

### 2015-08-06 related.tex
Complete pass.

Prior works on static race checking also include the following. Choi et al.~\cite{choi2002efficient} combine static analysis and runtime access caching to speed up dynamic race detection. Kahlon et al.~\cite{kahlon2007fast} use a divide-and-conquer algorithm that partitions all pointers of a program that do not alias in disjoint sets to achieve scalability; more recently, they used abstract interpretation to achieve a sound partial-order reduction on the set of thread interleavings and statically reduce the number of false race warnings~\cite{kahlon2009semantic}. Das et al.~\cite{das2015section} employ inter-procedural alias analysis and verifiable user annotations to split programs into disjoint sections, based on non-communicating accesses of shared data, and eliminate redundant checks during dynamic race detection. \whoop uses symbolic lockset analysis, which involves generating verification conditions and discharging them to a theorem prover, and then employs \corral to filter out false races.

>>>>

\COULDCUT{Prior works on static race checking also include the following.} Choi et al.~\cite{choi2002efficient} combine static analysis and runtime access caching to speed up dynamic race detection. Kahlon et al.~\cite{kahlon2007fast} use a divide-and-conquer algorithm that partitions all pointers of a program that do not alias in disjoint sets to achieve scalability; more recently, they used abstract interpretation to achieve a sound partial-order reduction on the set of thread interleavings and statically reduce the number of false race warnings~\cite{kahlon2009semantic}. Das et al.~\cite{das2015section} employ inter-procedural alias analysis and verifiable user annotations to split programs into disjoint sections, based on non-communicating accesses of shared data, and eliminate redundant checks during dynamic race detection. \whoop uses symbolic lockset analysis, which involves generating verification conditions and discharging them to a theorem prover, and then employs \corral to filter out false races.
